\chapter{Conclusion}
\label{ch:conclusion}

This project built and validated a numerical framework for computing
charged particle orbits in prescribed electromagnetic fields, then
applied it to study particle dynamics in the magnetic dipole field
across a range of conditions.

The solver, based on the adaptive RK45 scheme, was verified through a
hierarchy of test cases spanning circular gyromotion, helical motion,
$\mathbf{E} \times \mathbf{B}$ drift, gradient-$B$ drift, and the
dipole field itself. Numerically computed trajectories and drift
velocities were compared quantitatively against analytic predictions in
each case, with kinetic energy conservation monitored throughout as an
independent check. Relative energy errors stayed below $10^{-6}$ in
every test, confirming that the chosen tolerances are appropriate.

The validated solver was applied to full orbit integration in the dipole
field, demonstrating bounce motion between mirror points and approximate
conservation of the magnetic moment $\mu$. Numerically measured bounce
periods agreed with analytic values from the standard bounce-period
integral to within 1\,\%, consistent with the accuracy of the underlying
adiabatic approximation. A guiding-centre integrator was run alongside
the full Lorentz solver and shown to reproduce the slow bounce dynamics
faithfully across approximately 400 gyrations when the adiabatic
conditions are well satisfied, with the growing positional offset
attributable to secular phase drift in the non-symplectic RK45
integrator rather than breakdown of the approximation.

The framework was also applied in physical SI units, simulating a
10\,keV electron at $L = 3$ in Earth's dipole field. The numerically
computed mirror latitude agreed with the analytic prediction to within
0.02\,\%, and the gyroperiod (32\,$\mu$s), gyroradius (0.2\,km), and
bounce period (1.8\,s) all matched the expected values for this energy
and $L$-shell. An extension to tilted dipole fields showed that
displacing the magnetic equator from the geographic equatorial plane
produces asymmetric bounce motion in geographic coordinates, with the
degree of asymmetry increasing with tilt angle. Simulations at
$47^\circ$ and $59^\circ$ tilt reproduced the field geometries of
Neptune and Uranus respectively, confirming that the mirror mechanism
operates correctly regardless of the orientation of the dipole axis.

Adding a corotation electric field demonstrated the
$\mathbf{E} \times \mathbf{B}$ drift in the dipole geometry, recovering
the input rotation rate to within 96.6\,\%. The most complete
simulation combined a rotating tilted dipole with corotation, producing
time-dependent fields and reproducing all three adiabatic motions
(gyration, bounce, drift) simultaneously. The modulation of the bounce
amplitude by the rotating field geometry provides direct evidence that
the solver correctly couples the particle dynamics to the time-dependent
field.

The modular design of the implementation means that adding a new field
configuration requires only a new factory function in
\texttt{fields.py}, with no changes to the solver.

\section*{Future Work}

Several natural extensions would build on the framework developed here.

\begin{itemize}
  \item \textbf{Symplectic integrator.} Replacing the RK45 scheme with a
    symplectic particle pusher such as the Boris algorithm would suppress
    the secular phase drift observed in long integrations, enabling
    accurate tracking over thousands of bounce periods without energy or
    $\mu$ drift.

  \item \textbf{Relativistic velocities.} The current solver assumes
    $|\mathbf{v}| \ll c$. Extending the equations of motion to include
    the relativistic Lorentz factor $\gamma = (1 - v^2/c^2)^{-1/2}$
    would allow the framework to model higher-energy particles in
    radiation belts, where relativistic corrections become significant
    above approximately 100\,keV for electrons.

  \item \textbf{Non-dipolar field models.} Real planetary magnetospheres
    include higher-order multipole components, magnetopause currents, and
    a magnetotail. Incorporating empirical field models such as the
    Tsyganenko models for Earth would test the solver in more realistic
    geometries where analytic orbit theory no longer applies.

  \item \textbf{Time-dependent perturbations.} Adding oscillatory
    electric or magnetic field perturbations would enable the study of
    wave--particle interactions, a key mechanism for particle
    energisation and pitch-angle scattering in radiation belts.

  \item \textbf{Drift-shell splitting.} In a non-axisymmetric field,
    particles on the same $L$-shell but with different pitch angles
    drift onto different shells. This effect is absent in a pure dipole
    but would emerge naturally in a tilted or offset dipole with
    higher-order terms.
\end{itemize}
